---
title: 'P142-1 qPCR data - Re-analysing older dataset'
author: "Samuel O Skinner"
date: "March 29, 2017"
output:
  html_document:
    number_sections: yes
---


```{r global_opts, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=4, fig.align='center',
                      echo=TRUE, warning=FALSE, message=FALSE,
                      cache=FALSE, autodep=TRUE)
# knitr::opts_knit$set(root.dir = "..")

## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 1, digits = 5)
```

# Setup

## Prepare workspace


``` {r setup}

# R code to start with Fluidigm Ct data and return delta delta Ct values
path<-"~/Box Sync/Projects/Fluidigm_qPCR_analysis/"
directory<-"P142-2_Wambre_IL33_impact"

workdir<-paste(path, directory, "/", sep="")

library(combinat)
library(knitr)
library(shiny)
library(gplots)
library(ggplot2)
library(reshape2)
library(cowplot)
library(sva)

# read in scripts
source(file=paste(workdir,"scripts/","Fluidigm_R_Functions.R",sep=""))
source(file=paste(workdir,"scripts/","Fluidigm_R_HKgenes_Functions.R",sep=""))


``` 

## Load the data files and check for ID inconsistencies

``` {r read in parameters, echo=FALSE}

# read in all data files stored in Data directory - do not know how many there will be 
fileNames<-list.files(path=paste(workdir,"DataSubset1/",sep=""))
fileNames<-reorderFileNames(fileNames)

# read in data
dataList<-list()
for (i in 1:length(fileNames))
{
	dataList[[i]]<-readFluidData(paste(workdir,"DataSubset1/",fileNames[i],sep=""))
}
names(dataList)<-fileNames


# now check chamber IDs to make sure samples and genes match across plates
# if they do not match, then stop processing - need to fix the input data
anyError <- checkSampleGeneChamberIDs(dataList)
print(paste("Errors detected:", anyError))


# convert into matrix of rows (assays/genes) and columns (samples represented by chamber IDs)
ctValList<-list()
ctCallList<-list()
for (i in 1:length(dataList))
{
	ctValList[[i]]<-changeFluidFormat(dataList[[i]], colName="Ct.Value")
	ctCallList[[i]]<-changeFluidFormat(dataList[[i]], colName="Ct.Call")
}

# Extract and combine data from multiple plates
retList<-combineDataFromMultiplePlates(dataList, ctValList, ctCallList)

```




## Define the analysis parameters
``` {r define parameters, echo=FALSE}

# Define parameters
parameters <- data.frame(Key = c("PALX",
                                 "floor",
                                 "HKgenes",
                                 "RefSamples",
                                 "NegSamples"),
                         Value = c("0.20",
                                   "NA",
                                   "GAPDH GUSB HSP90",
                                   "Ambion 1ul of 1/4x;BioChain 1ul of 1/4x",
                                   "NTC"))

kable(parameters, row.names = FALSE)

```


# Preliminary analysis

## Visualize the raw data
``` {r vis raw, echo=FALSE, fig.height=20, fig.width=10}


raw <- retList$ctValDataForPalx
class(raw) <- "numeric"
raw[raw > 100] <- NA

colnames(raw) <- determineSampleNamesFromChamberIDInDataList(dataList, colnames(raw), retList$maxAssay)
rownames(raw) <- convertWellToGeneName(rownames(raw), dataList)[,2]

coords <- which(retList$ctCallDataForPalx == "Fail", arr.ind = TRUE)
coords[,2] <- 96*length(fileNames) + 1 - coords[,2]

# Plotting paramters
my_palette <- colorRampPalette(c("green", "white", "red"))(n = 299)

# lmat is the layout matrix, lwid defines column widths, lhei row heights
lmat <- rbind(c(0, 4, 2), c(0, 1, 3))
lwid <- c(0.5, 10, 2)
lhei <- c(0.75, 5)

par(mar = c(1, 1, 1, 1))
heatmap.2(t(raw),
          trace="none",
          dendrogram = "none",
          col=my_palette,
          na.color = "grey2",
          Rowv=FALSE,
          Colv=FALSE,
          add.expr=points(coords, 
                          col='#333333', bg='#333333', 
                          cex=0.5, pch=22),
          lmat = lmat, lwid = lwid, lhei = lhei, 
          #(bottom.margin, right.margin)
          margins = c(5, 2),
          denscol = "black",
          key.title = NA, key.ylab = NA, key.xlab = "Ct values",
          #( "bottom.margin", "left.margin", "top.margin", "right.margin" )
          key.par = list(mar = c(4, 4, 2, 4)))


```

Black in this plot represents reactions that did not produce a Ct value, where likely there was no template (in the case of a negative control, this would be a positive result). These are categorized as failed reactions by the fluidigm analysis packages.
  
The grey squares in this plot represent reactions labeled "Fail".  While most failed reactions did not produce a Ct value, notice that some reactions with Ct values have failed as well. These failures could be caused by aberrant/unexpected PCR products or contamination.
  
  


``` {r sep negs, echo=FALSE}

# Calculate the percentage present (or Call == Pass) for samples
SamPassMeans <- apply(retList$ctCallDataForPalx, 2, function(x) {mean(x=="Pass")})

plot_sam <- ggplot(data.frame(x=SamPassMeans),
       aes(x=x)) + 
  geom_line(stat = "density") +
  labs(x="FP", title="Fraction present for all samples")


# Calculate the percentage present (or Call == Pass) for genes (assays)
GenePassMeans <- apply(retList$ctCallDataForPalx, 1, function(x) {mean(x=="Pass")})

plot_gene <- ggplot(data.frame(x=GenePassMeans),
       aes(x=x)) + 
  geom_line(stat = "density") +
  labs(x="FP", title="Fraction present for all genes")

plot_grid(plot_sam, plot_gene)


```




## Analysis of the negative samples


```{r raw neg data, echo=FALSE, fig.height=4, fig.width=10}

# Negative sample root name
negNames <- strsplit(returnParameter(parameters, "NegSamples", "character"), split=";")[[1]]

# Identify the negative samples
sample_names <- determineSampleNamesFromChamberIDInDataList(dataList, colnames(retList$ctValDataForPalx), retList$maxAssay)
negList <- colnames(retList$ctValDataForPalx)[which(sample_names %in% negNames)]




rawValData  <- retList$ctValDataForPalx
rawCallData <- retList$ctCallDataForPalx
class(rawValData) <- "numeric"
rawValData[rawValData > 100] <- NA

negrawValData  <- rawValData[,colnames(rawValData) %in% negList]
negrawCallData <- rawCallData[,colnames(rawCallData) %in% negList]

colnames(negrawValData) <- determineSampleNamesFromChamberIDInDataList(dataList, colnames(negrawValData), retList$maxAssay)
rownames(negrawValData) <- convertWellToGeneName(rownames(negrawValData), dataList)[,2]



coords <- which(negrawCallData == "Fail", arr.ind = TRUE)
coords[,2] <- dim(negrawCallData)[2] + 1 - coords[,2]

# Plotting paramters
my_palette <- colorRampPalette(c("green", "white", "red"))(n = 299)

# lmat is the layout matrix, lwid defines column widths, lhei row heights
lmat <- rbind(c(0, 4, 2), c(0, 1, 3))
lwid <- c(0.5, 10, 2)
lhei <- c(1, 2)

par(mar = c(1, 1, 1, 1))
heatmap.2(t(negrawValData),
          trace="none",
          dendrogram = "none",
          col=my_palette,
          na.color = "grey2",
          Rowv=FALSE,
          Colv=FALSE,
          add.expr=points(coords, 
                          col='#333333', bg='#333333', 
                          cex=0.5, pch=22),
          lmat = lmat, lwid = lwid, lhei = lhei, 
          #(bottom.margin, right.margin)
          margins = c(5, 2),
          denscol = "black",
          key.title = NA, key.ylab = NA, key.xlab = "Ct values",
          #( "bottom.margin", "left.margin", "top.margin", "right.margin" )
          key.par = list(mar = c(4, 4, 2, 4)))

```




```{r negative samples, echo=FALSE}



# Calculate the percentage present for each sample (or Call == Pass)
negpassMeans_sam <- apply(negrawCallData, 2, function(x) {mean(x=="Pass")})
negpassNums_sam  <- apply(negrawCallData, 2, function(x) {sum(x=="Pass")})

# Calculate the mean Ct value for those that passed
negValPassNums_sam  <- apply(negrawValData, 2, function(x) { sum(as.numeric(x)<998, na.rm = T)})
negValPassMeans_sam <- apply(negrawValData, 2, function(x) {mean(as.numeric(x)[as.numeric(x)<998], na.rm = T)})
negValPassSds_sam   <- apply(negrawValData, 2, function(x) {  sd(as.numeric(x)[as.numeric(x)<998], na.rm = T)})


# Display the negative samples
if (length(negList)>0)
{
	negDisplay<-cbind(negList, 
	                  determineSampleNamesFromChamberIDInDataList(dataList, negList, retList$maxAssay),
	                  round(negpassMeans_sam, digits=3),
	                  round(negpassNums_sam, digits=3),
	                  round(negValPassNums_sam, digit=3),
	                  round(negValPassMeans_sam, digits=3),
	                  round(negValPassSds_sam, digits=3))
	colnames(negDisplay)<-c("ChamberID", "SampleName", "FracPass", "NumPass", "NumCt", "mean(Ct)", "std(Ct)")
}

kable(negDisplay, row.names=FALSE, caption = "No template controls")





# Calculate the percentage present for each assay (or Call == Pass)
negpassMeans_ass <- apply(negrawCallData, 1, function(x) {mean(x=="Pass")})
negpassNums_ass  <- apply(negrawCallData, 1, function(x) {sum(x=="Pass")})

# Calculate the mean Ct value for those that passed
negValPassNums_ass  <- apply(negrawValData, 1, function(x) { sum(as.numeric(x)<998, na.rm = T)})
negValPassMeans_ass <- apply(negrawValData, 1, function(x) {mean(as.numeric(x)[as.numeric(x)<998], na.rm = T)})
negValPassSds_ass   <- apply(negrawValData, 1, function(x) {  sd(as.numeric(x)[as.numeric(x)<998], na.rm = T)})

# Display the genes that appear in the negative samples
if (length(rownames(negrawCallData))>0)
{
	negDisplay<-cbind(convertWellToGeneName(rownames(negrawCallData), dataList),
	                  round(negpassMeans_ass, digits=3),
	                  round(negpassNums_ass, digits=3),
	                  round(negValPassNums_ass, digits=3),
	                  round(negValPassMeans_ass, digits=3),
	                  round(negValPassSds_ass, digits=3))
	colnames(negDisplay)<-c("ChamberID", "SampleName", "FracPass", "Numpass", "NumCt", "mean(Ct)", "std(Ct)")
}

kable(negDisplay, row.names=FALSE, caption = "Ct per gene in no template controls")



```


## Analysis of the reference samples

```{r analysis of reference samples, echo=FALSE}

rawValData  <- retList$ctValDataForPalx
rawCallData <- retList$ctCallDataForPalx
class(rawValData) <- "numeric"
rawValData[rawValData > 100] <- NA



# Reference sample root name
refNames <- strsplit(returnParameter(parameters, "RefSamples", "character"), ";")[[1]]

# Identify the reference samples
refList <- colnames(retList$ctValDataForPalx)[which(sample_names %in% refNames)]


refrawValData  <- rawValData[,colnames(rawValData) %in% refList]
refrawCallData <- rawCallData[,colnames(rawCallData) %in% refList]

colnames(refrawValData) <- determineSampleNamesFromChamberIDInDataList(dataList, colnames(refrawValData), retList$maxAssay)
rownames(refrawValData) <- convertWellToGeneName(rownames(refrawValData), dataList)[,2]



coords <- which(refrawCallData == "Fail", arr.ind = TRUE)
coords[,2] <- dim(refrawCallData)[2] + 1 - coords[,2]

# Plotting paramters
my_palette <- colorRampPalette(c("green", "white", "red"))(n = 299)

# lmat is the layout matrix, lwid defines column widths, lhei row heights
lmat <- rbind(c(0, 4, 2), c(0, 1, 3))
lwid <- c(0.5, 10, 2)
lhei <- c(1, 2)

par(mar = c(1, 1, 1, 1))
heatmap.2(t(refrawValData),
          trace="none",
          dendrogram = "none",
          col=my_palette,
          na.color = "grey2",
          Rowv=FALSE,
          Colv=FALSE,
          add.expr=points(coords, 
                          col='#333333', bg='#333333', 
                          cex=0.5, pch=22),
          lmat = lmat, lwid = lwid, lhei = lhei, 
          #(bottom.margin, right.margin)
          margins = c(5, 2),
          denscol = "black",
          key.title = NA, key.ylab = NA, key.xlab = "Ct values",
          #( "bottom.margin", "left.margin", "top.margin", "right.margin" )
          key.par = list(mar = c(4, 4, 2, 4)))


```





```{r reference stats, echo=FALSE}


# Calculate the percentage present for each sample (or Call == Pass)
refpassMeans_sam <- apply(refrawCallData, 2, function(x) {mean(x=="Pass")})
refpassNums_sam  <- apply(refrawCallData, 2, function(x) {sum(x=="Pass")})

# Calculate the mean Ct value for those that passed
refValPassNums_sam  <- apply(refrawValData, 2, function(x) { sum(as.numeric(x)<998, na.rm = T)})
refValPassMeans_sam <- apply(refrawValData, 2, function(x) {mean(as.numeric(x)[as.numeric(x)<998], na.rm = T)})
refValPassSds_sam   <- apply(refrawValData, 2, function(x) {  sd(as.numeric(x)[as.numeric(x)<998], na.rm = T)})


# Display the reference samples
if (length(refList)>0)
{
	refDisplay<-cbind(refList, 
	                  determineSampleNamesFromChamberIDInDataList(dataList, refList, retList$maxAssay),
	                  round(refpassMeans_sam, digits=3),
	                  round(refpassNums_sam, digits=3),
	                  round(refValPassNums_sam, digit=3),
	                  round(refValPassMeans_sam, digits=3),
	                  round(refValPassSds_sam, digits=3))
	colnames(refDisplay)<-c("ChamberID", "SampleName", "FracPass", "NumPass", "NumCt", "mean(Ct)", "std(Ct)")
}

kable(refDisplay, row.names=FALSE, caption = "Reference samples")





# Calculate the percentage present for each assay (or Call == Pass)
refpassMeans_ass <- apply(refrawCallData, 1, function(x) {mean(x=="Pass")})
refpassNums_ass  <- apply(refrawCallData, 1, function(x) {sum(x=="Pass")})

# Calculate the mean Ct value for those that passed
refValPassNums_ass  <- apply(refrawValData, 1, function(x) { sum(as.numeric(x)<998, na.rm = T)})
refValPassMeans_ass <- apply(refrawValData, 1, function(x) {mean(as.numeric(x)[as.numeric(x)<998], na.rm = T)})
refValPassSds_ass   <- apply(refrawValData, 1, function(x) {  sd(as.numeric(x)[as.numeric(x)<998], na.rm = T)})

# Display the genes that appear in the reference samples
if (length(rownames(refrawCallData))>0)
{
	refDisplay<-cbind(convertWellToGeneName(rownames(refrawCallData), dataList),
	                  round(refpassMeans_ass, digits=3),
	                  round(refpassNums_ass, digits=3),
	                  round(refValPassNums_ass, digits=3),
	                  round(refValPassMeans_ass, digits=3),
	                  round(refValPassSds_ass, digits=3))
	colnames(refDisplay)<-c("ChamberID", "SampleName", "FracPass", "Numpass", "NumCt", "mean(Ct)", "std(Ct)")
}

kable(refDisplay, row.names=FALSE, caption = "Ct per gene in no template controls")


```






## Analysis of the housekeeping genes

```{r analysis of housekeeping genes, echo=FALSE}

rawValData  <- retList$ctValDataForPalx
rawCallData <- retList$ctCallDataForPalx
class(rawValData) <- "numeric"
rawValData[rawValData > 100] <- NA



# Housekeeping gene names
HKgeneNames <- strsplit(returnParameter(parameters, "HKgenes", "character"), " ")[[1]]

# Identify the HK genes' assay numbers
HKgeneList <- rownames(rawValData)[which(convertWellToGeneName(rownames(rawValData), dataList)[,2] %in% HKgeneNames)]


HKrawValData  <- rawValData[rownames(rawValData) %in% HKgeneList,]
HKrawCallData <- rawCallData[rownames(rawCallData) %in% HKgeneList,]

colnames(HKrawValData) <- determineSampleNamesFromChamberIDInDataList(dataList, colnames(HKrawValData), retList$maxAssay)
rownames(HKrawValData) <- convertWellToGeneName(rownames(HKrawValData), dataList)[,2]


coords <- which(t(HKrawCallData) == "Fail", arr.ind = TRUE)
coords[,2] <- dim(HKrawCallData)[1] + 1 - coords[,2]

# Plotting paramters
my_palette <- colorRampPalette(c("green", "white", "red"))(n = 299)

# lmat is the layout matrix, lwid defines column widths, lhei row heights
lmat <- rbind(c(0, 4, 2), c(0, 1, 3))
lwid <- c(0.5, 10, 2)
lhei <- c(1, 2)

par(mar = c(1, 1, 1, 1))
heatmap.2((HKrawValData),
          trace="none",
          dendrogram = "none",
          col=my_palette,
          na.color = "grey2",
          Rowv=FALSE,
          Colv=FALSE,
          add.expr=points(coords, 
                          col='#333333', bg='#333333', 
                          cex=0.5, pch=22),
          lmat = lmat, lwid = lwid, lhei = lhei, 
          #(bottom.margin, right.margin)
          margins = c(5, 2),
          denscol = "black",
          key.title = NA, key.ylab = NA, key.xlab = "Ct values",
          #( "bottom.margin", "left.margin", "top.margin", "right.margin" )
          key.par = list(mar = c(4, 4, 2, 4)))


```





```{r housekeeping stats, echo=FALSE}
# 
# 
# # Calculate the percentage present for each sample (or Call == Pass)
# HKpassMeans_sam <- apply(HKrawCallData, 2, function(x) {mean(x=="Pass")})
# HKpassNums_sam  <- apply(HKrawCallData, 2, function(x) {sum(x=="Pass")})
# 
# # Calculate the mean Ct value for those that passed
# HKValPassNums_sam  <- apply(HKrawValData, 2, function(x) { sum(as.numeric(x)<998, na.rm = T)})
# HKValPassMeans_sam <- apply(HKrawValData, 2, function(x) {mean(as.numeric(x)[as.numeric(x)<998], na.rm = T)})
# HKValPassSds_sam   <- apply(HKrawValData, 2, function(x) {  sd(as.numeric(x)[as.numeric(x)<998], na.rm = T)})
# 
# 
# # Display the HKerence samples
# if (length(refList)>0)
# {
# 	HKDisplay<-cbind(refList, 
# 	                  determineSampleNamesFromChamberIDInDataList(dataList, refList, retList$maxAssay),
# 	                  round(HKpassMeans_sam, digits=3),
# 	                  round(HKpassNums_sam, digits=3),
# 	                  round(HKValPassNums_sam, digit=3),
# 	                  round(HKValPassMeans_sam, digits=3),
# 	                  round(HKValPassSds_sam, digits=3))
# 	colnames(HKDisplay)<-c("ChamberID", "SampleName", "FracPass", "NumPass", "NumCt", "mean(Ct)", "std(Ct)")
# }
# 
# kable(HKDisplay, row.names=FALSE, caption = "Housekeeping genes")
# 
# 
# 
# 
# 
# # Calculate the percentage present for each assay (or Call == Pass)
# HKpassMeans_ass <- apply(HKrawCallData, 1, function(x) {mean(x=="Pass")})
# HKpassNums_ass  <- apply(HKrawCallData, 1, function(x) {sum(x=="Pass")})
# 
# # Calculate the mean Ct value for those that passed
# HKValPassNums_ass  <- apply(HKrawValData, 1, function(x) { sum(as.numeric(x)<998, na.rm = T)})
# HKValPassMeans_ass <- apply(HKrawValData, 1, function(x) {mean(as.numeric(x)[as.numeric(x)<998], na.rm = T)})
# HKValPassSds_ass   <- apply(HKrawValData, 1, function(x) {  sd(as.numeric(x)[as.numeric(x)<998], na.rm = T)})
# 
# # Display the genes that appear in the reference samples
# if (length(rownames(HKrawCallData))>0)
# {
# 	HKDisplay<-cbind(convertWellToGeneName(rownames(HKrawCallData), dataList),
# 	                  round(HKpassMeans_ass, digits=3),
# 	                  round(HKpassNums_ass, digits=3),
# 	                  round(HKValPassNums_ass, digits=3),
# 	                  round(HKValPassMeans_ass, digits=3),
# 	                  round(HKValPassSds_ass, digits=3))
# 	colnames(HKDisplay)<-c("ChamberID", "SampleName", "FracPass", "Numpass", "NumCt", "mean(Ct)", "std(Ct)")
# }
# 
# kable(HKDisplay, row.names=FALSE, caption = "Ct per gene in no template controls")


```

